PLAE enables quick and easy searching across one million ocular transcriptomes

### 2021-11-13 manuscript_00.Rmd
text updates

---
title: 'PLAE enables quick and easy searching across one million ocular transcriptomes'
author:
  - Vinay Swamy:
      institute: bg
  - Zachary Batz:
      institute: nnrl
  - Robert Hufnagel:
      institute: mgog
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We wrote the Shiny web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (sc Eye in a Disk, scEiaD) to display scEiaD. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes., XX cell types, XX ocular tissues, XX body tissues. As a test of the value of this unified pan-eye dataset, we show XX. CONCLUSION: The PLAE v0.80 web app serves the pan-ocular and body dataset, scEiaD. This offers the eye community a powerful and quick means to test hypotheses on gene and transcript expression across 54 body and 19 eye tissues. "
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

library(pool)
library(RSQLite)
meta_filter_v0 <- fst::read_fst("~/data/scEiaD/2021_03_17_meta_filter.fst")
# remove eric bharti 2D/3D RPE
meta_filter <- fst::read_fst("~/data/scEiaD_v2/2021_10_22_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D')

@@

# sup fig 1 
Effect of decontx for Rho between scEiaD v2 and scEiaD v1

<!-- ```{r, fig.height=5, fig.width=3} -->
<!-- # Distribution of ML Retina Cell Types Between Mouse and Human in Pan-Retina Studies -->
<!-- Drop this figure? -->
<!-- Yes -->

@@

<!-- meta_filter %>%  -->
<!--   filter(organism %in% c('Mus musculus','Homo sapiens'),  -->
<!--          Stage == 'Mature',  -->
<!--          study_accession %in% pan_retina_studies,  -->
<!--          grepl('Amacr|Bipol|Cone|Epithel|Endoth|Horizon|Muller|Pericy|Ganglion|Rod', CellType_predict)) %>%  -->
<!--   group_by(study_accession, CellType_predict) %>%  -->
<!--   summarise(Count = n()) %>% mutate(Percentage = 100 * (Count / sum(Count)))  %>%  -->
<!--   left_join(meta_filter %>% select(study_accession, Platform, organism) %>% unique()) %>%  -->
<!--   ggplot(aes(x=organism, y=Percentage)) +  -->
<!--   geom_boxplot() +  -->
<!--   coord_flip() +  -->
<!--   ggbeeswarm::geom_quasirandom(aes(color=study_accession)) +  -->
<!--   facet_wrap(~CellType_predict, ncol = 1) +  -->
<!--   cowplot::theme_cowplot() -->
<!-- ``` -->
# Introduction
## Many published ocular atlases

## Pan Atlases contain limited retina cell types or lack integration

Tabula Sapiens, MOCA (?), BBrowser (\$\$\$), https://cells.ucsc.edu

@@

## Updated the scEiaD backend
Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types. We updated the scEiaD resource for the PLAE v0.80 web app in five important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset. Third, we added a human pan-body reference scRNA dataset. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX). Fourth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450. Fifth, we carried and harmonnized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.

@@

# Comparison of PLAE to "pan" web apps with retina data
```{r}
feature_table <- cbind(c('Spectacle', 'No' ,'No', ),
                       c('UCSC Cell Browser', 'Yes', 'No', 'https://cells.ucsc.edu' ),
                       c('Tabula Sapiens', 'No', 'N/A', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
                       c('PLAE','Yes','Yes', 'plae.nei.nih.gov'),
                       c('PanglaoDB', 'Yes', 'No'),
                       c('Protein Atlas', 'Yes','No', 'https://www.proteinatlas.org/humanproteome/celltype'))
colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'URL')
```

>>>>

---
title: 'PLAE enables quick and easy searching across one million ocular transcriptomes'
author:
  - Vinay Swamy:
      institute: bg
  - Zachary Batz:
      institute: nnrl
  - Robert Hufnagel:
      institute: mgog
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health
```{r}

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We wrote the Shiny web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (single cell Eye in a Disk, scEiaD) to asdfasdf. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes., XX cell types, XX ocular tissues, XX body tissues. As a test of the value of this unified pan-eye dataset, we show XX. CONCLUSION: The PLAE v0.80 web app serves the pan-ocular and body dataset, scEiaD. This offers the eye community a powerful and quick means to test hypotheses on gene and transcript expression across 54 body and 19 eye tissues. "
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

library(pool)
library(RSQLite)
meta_filter_v0 <- fst::read_fst("~/data/scEiaD/2021_03_17_meta_filter.fst") %>% 
  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType))
# remove eric bharti 2D/3D RPE
meta_filter <- fst::read_fst("/Volumes/McGaughey_S/scEiaD_v3/2021_11_09_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType))

@@

# Sup fig 1 
DecontX *in silico* ambient RNA contamination tool substantially removes Rhodopsin expression in non-rod cells while retaining high expression rods. First shown (a) is Rhodopsin expression across the scEiaD v0 dataset, without DecontX optimization. Note Rhodopsin expression is noticeable in many cell types beyond the rods. Next (b) is the scEiaD v1 dataset with DecontX optimization. Rhodopsin is nearly exclusively expressed in labelled rod cells. Despite the computational ambient RNA removal, Rhodopsin expression remains high in the Rods.

```{r, fig.width=10, fig.height=12}
source('src/make_exp_plot.R')
scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  '~/data/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite')
scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  '/Volumes/McGaughey_S/scEiaD_v3/MOARTABLES__anthology_limmaFALSE___6000-counts-universe-batch-scVIprojection-10-15-0.1-50-20.sqlite')

input <- list()
input$exp_plot_groups <- c('organism')
input$exp_plot_facet <- c('CellType_predict')
input$exp_plot_genes <- c('RHO (ENSG00000163914)')
input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
input$exp_plot_col_num <- 16
input$exp_filter_min_cell_number <- 50

v0_rho <- make_exp_plot(input, scEiaD_2020_v00, meta_filter_v0)
v1_rho <-  make_exp_plot(input, scEiaD_2020_v01, meta_filter)
input$exp_filter_min_cell_number <- 1000
v1.1_rho <- make_exp_plot(input, scEiaD_2020_v01, meta_filter %>% filter(CellType_predict %in% unique(meta_filter_v0$CellType_predict)))
plot_grid(v0_rho, v1_rho, ncol = 1, labels = c('a','b'), rel_heights = c(1,1.5))
```
<!-- ```{r, fig.height=5, fig.width=3} -->
<!-- # Distribution of ML Retina Cell Types Between Mouse and Human in Pan-Retina Studies -->
<!-- Drop this figure? -->
<!-- Yes -->

@@

<!-- meta_filter %>%  -->
<!--   filter(organism %in% c('Mus musculus','Homo sapiens'),  -->
<!--          Stage == 'Mature',  -->
<!--          study_accession %in% pan_retina_studies,  -->
<!--          grepl('Amacr|Bipol|Cone|Epithel|Endoth|Horizon|Muller|Pericy|Ganglion|Rod', CellType_predict)) %>%  -->
<!--   group_by(study_accession, CellType_predict) %>%  -->
<!--   summarise(Count = n()) %>% mutate(Percentage = 100 * (Count / sum(Count)))  %>%  -->
<!--   left_join(meta_filter %>% select(study_accession, Platform, organism) %>% unique()) %>%  -->
<!--   ggplot(aes(x=organism, y=Percentage)) +  -->
<!--   geom_boxplot() +  -->
<!--   coord_flip() +  -->
<!--   ggbeeswarm::geom_quasirandom(aes(color=study_accession)) +  -->
<!--   facet_wrap(~CellType_predict, ncol = 1) +  -->
<!--   cowplot::theme_cowplot() -->
<!-- ``` -->

# Introduction

## Many published ocular atlases

## Several web atlases with ocular cells exist but have significant limitations

There are several web app resources for querying gene expression across ocular cell types (table 1).

@@

## Updates to the scEiaD backend
Version zero of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types. We updated the scEiaD resource for the PLAE v0.80 web app in five important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added human ocular outflow tract and cornea datasets. Third, we added a human pan-body reference scRNA dataset. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX). Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450. Sixth, we carried and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.

@@

# Comparison of PLAE to "pan" scRNA web apps containing ocular data
```{r}
feature_table <- rbind(
  c('PLAE','Yes','Yes', 'Yes', '887', 'https://plae.nei.nih.gov'),
  c('Spectacle', 'No' ,'No', 'No', '87', 'https://singlecell-eye.org'),
  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
  c('PanglaoDB', 'Yes', 'No', 'Yes', '47', 'https://panglaodb.se'),
  c('Protein Atlas', 'Yes*','No', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
  as_tibble()
colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'Ocular and non-Ocular Data', 'Ocular Cell Count\n(thousands)', 'URL')
feature_table %>% flextable()
# as of 2021 10 25
```
